---
title: "Ecotourism Tutorial"
subtitle: "Joining occurrence, tourism, and weather data"
author: "GSoC 2026 Contributor"
quarto-required: ">=1.8.0"
format:
    unilur-html:
        output-file: tutorial.html
        embed-resources: true
        css: "../assets/tutorial.css"
    unilur-html+solution:
        output-file: tutorialsol.html
        embed-resources: true
        css: "../assets/tutorial.css"
        show-solution: true
---

```{r include=FALSE}
#| echo: false
library(tidyverse)
library(lubridate)
library(ecotourism)
library(kableExtra)
library(conflicted)

# Set up chunk for all slides
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
options(
  digits = 2,
  width = 80,
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  ggplot2.discrete.colour = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442"),
  ggplot2.discrete.fill = c("#D55E00", "#0072B2", "#009E73", "#CC79A7", "#E69F00", "#56B4E9", "#F0E442")
)
theme_set(theme_bw(base_size = 14) +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title.position = "plot",
     plot.title = element_text(size = 24),
     panel.background = element_rect(fill = 'transparent', colour = NA),
     legend.background = element_rect(fill = 'transparent', colour = NA),
     legend.key = element_rect(fill = 'transparent', colour = NA)
   )
)

conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::slice)

tourism_total <- tourism_quarterly |>
  group_by(ws_id, year, quarter) |>
  summarise(total_trips = sum(trips, na.rm = TRUE), .groups = "drop")
```

## 🎯 Objectives

Practice joining wildlife occurrence records with tourism and weather data from the **ecotourism** package, and explore how visitor numbers relate to wildlife sightings across different regions, organisms, and months.

## 🔧 Preparation

- Install the **ecotourism** package by running `remotes::install_github("vahdatjavad/ecotourism")` in the Console.
- Complete the weekly quiz, before the deadline!
- Open your RStudio Project for this unit, (the one you created in week 1). Create a `.qmd` document for this week's activities.

## 📥 Exercises

The **ecotourism** package contains wildlife occurrence records and quarterly tourism trip counts for Australia, linked by `ws_id`. In the setup chunk, `tourism_total` has been created by summing trips across purposes per `ws_id`, `year`, and `quarter`.

```{r}
#| echo: false
desc <- tibble(
  Object = c("gouldian_finch", "manta_rays", "glowworms",
             "tourism_quarterly", "weather"),
  Description = c(
    "Gouldian Finch occurrence records (2014–2024)",
    "Manta Ray occurrence records (2014–2024)",
    "Glowworm occurrence records (2014–2024)",
    "Quarterly tourism trips by region and purpose",
    "Daily weather by station: temp, max, min, prcp linked by ws_id and date"
  )
)
kbl(desc, table.attr = 'data-quarto-disable-processing="true"') |>
  kable_styling(full_width = FALSE) |>
  column_spec(1, width = "4cm", border_right = TRUE) |>
  column_spec(2, width = "10cm")
```

**Question 1 (one region).** Join `gouldian_finch` with `tourism_total` and `ecotourism::weather`, keeping only sightings from **Western Australia** (`obs_state == "Western Australia"`). How many sightings are in the joined table? How many have a non-`NA` value for `max` (daily maximum temperature)? What does this tell you about the completeness of the weather data in the current package?

::: unilur-solution
```{r}
wa_joined <- gouldian_finch |>
  filter(obs_state == "Western Australia") |>
  mutate(quarter = lubridate::quarter(date)) |>
  left_join(tourism_total, by = c("ws_id", "year", "quarter")) |>
  left_join(
    ecotourism::weather |> dplyr::select(ws_id, date, temp, max, min, prcp),
    by = c("ws_id", "date")
  )

nrow(wa_joined)
sum(!is.na(wa_joined$max))
```

Many `max` values will be `NA` because the weather records in the current package version are incomplete — not all stations have full coverage. Completing the weather records across all Australian stations is one of the main goals of the GSoC 2026 project.
:::

**Question 2 (one organism).** Join `manta_rays` with `tourism_total` and `ecotourism::weather` across all regions and months. For each month, calculate the mean total tourism trips and the proportion of sighting days that have a non-`NA` `max` value. In which months is weather data available, and what does the pattern of missing weather tell you?

::: unilur-solution
```{r}
manta_joined <- manta_rays |>
  mutate(quarter = lubridate::quarter(date)) |>
  left_join(tourism_total, by = c("ws_id", "year", "quarter")) |>
  left_join(
    ecotourism::weather |> dplyr::select(ws_id, date, temp, max, min, prcp),
    by = c("ws_id", "date")
  )

manta_joined |>
  group_by(month) |>
  summarise(
    mean_trips  = mean(total_trips, na.rm = TRUE),
    n_sightings = n(),
    pct_weather = mean(!is.na(max)) * 100
  ) |>
  arrange(desc(mean_trips))
```

The `pct_weather` column shows which months have weather coverage for manta ray sighting locations. Where it is low or zero, weather records are missing for those stations — completing this coverage is a key goal of the GSoC 2026 project.
:::

**Question 3 (one month).** Join `glowworms` with `tourism_total` and `ecotourism::weather`, keeping only sightings from **April** (`month == 4`). Make a scatter plot of `total_trips` against `year`, colouring by `obs_state`. Add a note to the plot title indicating how many sighting days have weather data available. What would this plot look like once the weather data is complete?

::: unilur-solution
```{r}
glowworm_april <- glowworms |>
  filter(month == 4) |>
  mutate(quarter = lubridate::quarter(date)) |>
  left_join(tourism_total, by = c("ws_id", "year", "quarter")) |>
  left_join(
    ecotourism::weather |> dplyr::select(ws_id, date, temp, max, min, prcp),
    by = c("ws_id", "date")
  )

n_weather <- sum(!is.na(glowworm_april$max))

glowworm_april |>
  distinct(ws_id, year, obs_state, total_trips) |>
  ggplot(aes(x = year, y = total_trips, colour = obs_state)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = paste0("Tourism trips in glowworm regions — April",
                   " (weather available for ", n_weather, " sighting days)"),
    x      = "Year",
    y      = "Total quarterly trips",
    colour = "State"
  )
```

Once the GSoC update delivers complete weather records, `max` and `min` temperature could be overlaid on this plot to reveal whether cooler April nights correlate with higher glowworm tourism activity.
:::

## 👌 Finishing up

Make sure you say thanks and good-bye to your tutor. This is a time to also report what you enjoyed and what you found difficult.
